NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2019-10-31 Manuscript/manuscriptPGPE.tex
Update text

\begin{abstract}
% insert abstract here
  Primarily measured but also simulated NMR order parameters will be collected also for other than phophatidylcholine
  (these are discussed in NMRlipids I) headgroup. The information will be used to understand structural differences between 
  different lipid molecules in bilayers.
\end{abstract}

@@

PE and PG lipids are most common lipids in bacteria \cite{sohlenkamp16}.
Zwitterionic PE is the second most abundant glycerophospholipid in eukaryotic cells
and has been related to the diseases \cite{vance15,calzada16,patel17}.
Anionic PG lipids are less abundant, but is also proposed to be fundamental for terrestrial life \cite{furse17}.
PE and PG affect membrane protein functionality \cite{hariharan18} and bind to various proteins \cite{yeagle14}.
PE headgroup is also prone for negative membrane curvature and causes membrane fusion \cite{??}.
Therefore, the PE and PG headgroup structures play probably essential roles in 
many biological processes.

Structural details of lipid headgroups are mainly studied using NMR experiments, which
suggest that the glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, 
glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90},
and the headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. 
%Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
%In contrast to PC lipids (see \cite{botan15} and references therein).
Some attempts to resolve conformational ensembles from NMR for PC lipids have been made,
but lesser extend for other lipids \cite{??}.
Classical molecular dynamics simulations could potentially give such ensembles and therefore enable
the detailed studies of lipid headgroup behaviour in complex biomolecular systems, but current
force fields are not accurate enough to reproduce the correct conformational ensembles for PC and PS headgroups \cite{botan15,antila19}.
Several MD simulations of PE and PG lipids have been published especially in the context of modeling
inner membrane of Gram-negative bacteria \cite{??}, but glycerol backbone and headgroup structures have not
been evaluated against experiments.

Besides the structure, also ion binding may regulate biophysical activity of especially negatively
charged lipid headgroups \cite{??}. Monovalent cation (except Lithium) binding to zwitterionic PC headgroups is
very weak, while multivalent ion binding is stronger but still weak \cite{??}. The ion binding affinity
data for PE is more scarce \cite{??}, but large differences to PC would be surprising.
Negatively charged lipids are suggested to bear same cation binding constants than zwitterionic lipids, but
the amount of bound ions to negatively charged membranes would still be larger because
the concentration of cations in the vicinity of membranes would be higher  \cite{seelig90}.
%is stronger
%Based on the electrometer concept and other data is has been suggested that \cite{seelig90}
%\begin{displayquote}
%  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
%    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
%    (ii) the free Ca$^{2+}$
%    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
%    charge, either due to protein or to lipid; \\
%    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
%    of bound Ca$^{2+}$ at neutral and charged lipids; \\
%    (iv) the actual binding step can be described by a Langmuir
%    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
%    used to describe the chemical binding equilibrium.''}
%\end{displayquote}
On the other hand, anionic PS lipids are proposed chelate with calcium ions \cite{??}.
In simulations, the cation binding affinity to PC and PS membranes is typically overestimated \cite{catte16,antila19},
which can be improved by applying the ECC to the partial charges of the force fields \cite{melcr18,melcr19}.

>>>>

\begin{abstract}
% insert abstract here
Abstract
\end{abstract}

@@

PE and PG lipids are most common lipids in bacteria \cite{sohlenkamp16}.
Zwitterionic PE is the second most abundant glycerophospholipid in eukaryotic cells
and has been related to the diseases \cite{vance15,calzada16,patel17}.
Anionic PG lipids are less abundant, but is also proposed to be fundamental for terrestrial life \cite{furse17}.
PE and PG affect membrane protein functionality \cite{hariharan18} and bind to various proteins \cite{yeagle14}.
PE headgroup is also prone for negative membrane curvature and causes membrane fusion \cite{Chernomordik08,calzada16}.
Therefore, the PE and PG headgroup structures play probably essential roles in 
many biological processes.

Structural details of lipid headgroups are mainly studied using NMR experiments, which
suggest that the glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, 
glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90},
and the headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. 
%Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
%In contrast to PC lipids (see \cite{botan15} and references therein).
Some attempts to resolve conformational ensembles from NMR for PC and PE lipids have been made,
but lesser extend for PG or PS lipids \cite{seelig77c,davis83,Semchyschyn04}.
Classical molecular dynamics simulations could potentially give such ensembles and therefore enable
the detailed studies of lipid headgroup behaviour in complex biomolecular systems, but current
force fields are not accurate enough to reproduce the correct conformational ensembles for PC and PS headgroups \cite{botan15,antila19}.
Several MD simulations of PE and PG lipids have been published especially in the context of modeling
inner membrane of Gram-negative bacteria \cite{devries04,murzyn05,pedersen06,zhao07,gurtovenko08,zhao08,kukol09,tsai12,dickson12,venable13,dickson14,berglund15}
\todo{There may be some relevant publication missing from here},
but glycerol backbone and headgroup structures have not
been evaluated against experiments.

Besides the structure, also ion binding may regulate biophysical activity of especially negatively
charged lipid headgroups \cite{seelig90}. Monovalent cation (except Lithium) binding to zwitterionic PC and anionic PS headgroups is
very weak, while multivalent ion binding is stronger but still weak \cite{cevc90,tocanne90,roux90,catte16,antila19}. The ion binding affinity
data for PE is more scarce \cite{marra85}, but large differences to PC would be surprising.
Negatively charged lipids are suggested to bear same cation binding constants than zwitterionic lipids, but
the amount of bound ions to negatively charged membranes would still be larger because
the concentration of cations in the vicinity of membranes would be higher  \cite{seelig90}.
%is stronger
%Based on the electrometer concept and other data is has been suggested that \cite{seelig90}
%\begin{displayquote}
%  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
%    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
%    (ii) the free Ca$^{2+}$
%    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
%    charge, either due to protein or to lipid; \\
%    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
%    of bound Ca$^{2+}$ at neutral and charged lipids; \\
%    (iv) the actual binding step can be described by a Langmuir
%    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
%    used to describe the chemical binding equilibrium.''}
%\end{displayquote}
On the other hand, anionic PS lipids are proposed chelate with calcium ions \cite{hauser85,feigenson86,roux91}.
In simulations, the cation binding affinity to PC and PS membranes is typically overestimated \cite{catte16,antila19},
which can be improved by applying the ECC to the partial charges of the force fields \cite{melcr18,melcr19}.
